---
---

# Microarray Based Tumor Classification {-}

Microarray technologies allow scientists to quickly and accurately profile the
expression of tens of thousands of genes at a time at a fraction of the cost of
modern day sequencing protocols. Repositories and databases such as the Gene
Expression Omnibus (GEO) are home to tens of thousands of published and
unpublished microarray experiments, and millions of individual samples.
Microarray gene expression data are useful in a broad range of research tasks,
including disease diagnosis, drug discovery, and toxicological research.
Several companies have been born out of the application of this technology to
disease prediction such as the Oncotype Dx colon, breast, and prostate cancer
prediction assays or Agendia’s MammaPrint breast cancer test. With such a
tremendous amount of publicly available data and the translational impact of
modern microarray analyses, it is essential that every bioinformatics
researcher possess the knowledge and skill set to understand, analyze, and
interpret these data. This project will give you first-hand experience in
acquiring and analyzing a public microarray dataset.

This analysis will focus only on reproducing the results from the comparison
of the C3 and C4 tumor subtypes. The study was conducted in a two-phase design,
where an initial set of "discovery" samples was used to identify patterns
among the samples, and a separate set of "validation" samples was used to
test if the results from the discovery set were robust. For this analysis, we
have combined the discover and validation set samples into a single dataset
that you will use. There are 134 samples in total.

Upon the completion of project 1, students will be able to do the following:

* Read and understand the detailed computational methods of a relevant
  microarray analysis paper.
* Navigate the Gene Expression Omnibus (GEO) database and download publicly
  available microarray data.
* Upload data to a remote server (SCC).
* Compose and document original scripts using the R programming language.
* Normalize and quality control a large dataset of microarray samples.
* Employ noise filtering techniques to reduce data dimensionality.
* Perform data-driven analyses, such as hierarchical clustering, to discover
  novel relationships among samples in a given dataset.

Contents:

* [1. Read the paper & supplemental methods]
* [2. Data acquisition & transfer to remote server]
* [3. Data preprocessing & quality control]
* [4. Noise filtering & dimensionality reduction]
* [5. Hierarchical clustering & subtype discovery]
* [6. In-depth Analysis]
* [7. Discuss Your Findings]
* [Assignment Writeup]

## 1. Read the paper & supplemental methods {-}

[Marisa et al. Gene Expression Classification of Colon Cancer into Molecular
Subtypes: Characterization, Validation, and Prognostic Value. PLoS Medicine,
May 2013. PMID: 23700391](https://www.ncbi.nlm.nih.gov/pubmed/23700391)

## 2. Data acquisition & transfer to remote server {-}

**Lead Role: Data Curator**

Create a directory for your group under the following path:

```
/projectnb/bf528/users/
```

For example, you might create the directory `group_1`:

```
mkdir /projectnb/bf528/users/group_1
```

Then create a directory for this project under that directory:

```
mkdir /projectnb/bf528/users/group_1/project_1
```

Create directories under this directory for use in this project, and follow
this pattern for subsequent projects.This study is fairly large. To prevent the
download and storage of multiple copies of this dataset, we have downloaded all
but one of the samples and have deposited them in a central location on SCC
(`/project/bf528/project_1/data/GSE39582/CEL_files`). Your task will be to
download the missing sample (sample ID: GSM971958) and to upload it to your
`samples` directory. There are two popular approaches to getting files onto a
remote server. You can download the file to your personal computer and upload
it using an SFTP client of your choosing, or you can utilize the `wget`
utility on SCC to download the file directly to the file system.

1. Locate the paper’s repository on [GEO](http://www.ncbi.nlm.nih.gov/geo/) by
   searching for accession number GSE39582.
2. Identify the sample GSM971958 and download the .CEL.gz file to your SCC
   directory.
3. Since we are using relative paths in our analysis code to maintain
   portability, we would like to have all CEL files available in our
   `samples` directory.  However, we also would like to avoid having
   duplicates of files on our system as this is a waste of resources. We can
   make use of **symbolic links** to make a file *appear* to exist in a
   directory with the command `ln -s <source> <dest>`.  Use the `ln -s`
   command to create symbolic links of all of the other CEL files from the
   location above into your `samples` directory. Hint: the `ln` command can
   accept multiple `<source>` arguments when `<dest>` is a directory. This
   will create a symbolic link for every file in `<source>` in the `<dest>`
   directory.

## 3. Data preprocessing & quality control {-}

**Lead Role: Programmer**

Normalization of microarray data is necessary to ensure that differences in
intensities read by the scanner are due to differential gene expression and not
due to printing, hybridization, or scanning artifacts. Several methods exist
for normalizing microarray samples together. Many include quantile
normalization, log-median centering, or the use of control genes. Fortunately
for us, there are many researchers who devote their time to making tasks such
as this as easy and straightforward as possible. Here, you will need to write
an R script to normalize all of the microarrays together using the Robust
Multiarray Averaging (RMA) algorithm. Then, you will compute standard quality
control metrics on your normalized data and visualize the distribution of
samples using Principal Component Analysis (PCA).

1. There are two major repositories we will be leveraging during this course:
   CRAN and Bioconductor. As mentioned before, to download packages from CRAN,
   issue the command `install.packages("packageNameHere")` from within your R
   session. To install packages from Bioconductor, its necessary to first source
   the biocLite function on the Bioconductor servers:

   ```{r, eval=FALSE}
   if (!require("BiocManager", quietly = TRUE))
     install.packages("BiocManager")
   BiocManager::install(version = "3.14")
   ```

   Once this function has been sourced, you can use the command `BiocManager::install("packageNameHere")` to
   automatically download any Bioconductor package. For this assignment, you
   will need the `affy`, `affyPLM`, `sva`, `AnnotationDbi`, and
   `hgu133plus2.db` packages located in Bioconductor. Download and install
   these packages to SCC. For this assignment, we will be working in R. Issue
   the following commands to load the R module and install your packages. When
   prompted with a `yes or no [y/n]`, respond with yes::

   ```{r, eval=FALSE}
   module load R
   R --vanilla
   if (!require("BiocManager", quietly = TRUE))
       install.packages("BiocManager")
   BiocManager::install(version = "3.14")
   BiocManager::install(c("affy","affyPLM","sva","AnnotationDbi","hgu133plus2.db"))
   ```

2. Once installed, confirm that you can load each of these packages by opening
 a new R session and typing `library(packageNameHere)`. These commands should
 execute without error.
3. Read in the CEL files using the `ReadAffy()` function, then use the
 `rma()` function to normalize all of the CEL files together. If you are
 running into trouble here, try issuing the commands `?ReadAffy` or
 `?rma` within your R session. This will bring up the help files for these
 functions. For further help, check out the [affy vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/affy/inst/doc/affy.pdf)
 for a detailed walkthrough of the normalization process.
4. Using the Bioconductor package `affyPLM`, compute Relative Log Expression
   (RLE) and Normalized Unscaled Standard Error (NUSE) scores of the microarray
   samples. You will need to provide the output of your `ReadAffy()` call
   (*not* from `rma()`) from 3.3 with the `fitPLM` function, and provide
   `normalize=TRUE` and `background=TRUE` as additional arguments.
   Summarize these data by computing the median RLE and NUSE for each sample,
   then examine the distribution of the medians by plotting them in a
   histogram.
5. Use `ComBat` (`sva` package) to correct for batch effects. We have
   provided an annotation file for you which contains a host of clinical and
   batching annotation used by the authors for their analysis. This file can be
   found on SCC (`/project/bf528/project_1/doc/proj_metadata.csv`). Batch
   effects include both Center and RNA extraction method and have been merged
   into a single variable called `normalizationcombatbatch` in the annotation
   file.  Features of interest include both tumor and MMR status as per Marisa
   et. al.  and have been merged into a single variable called
   `normalizationcombatmod`. Use these two variables when running ComBat to
   correct for batch effects while preserving features of interest. Write out
   the expression data to a CSV file using the `exprs` and `write.csv`
   functions, where probesets are rows and samples are columns.
6. Perform Principal Component Analysis (PCA) on your normalized data using the
   `prcomp()` function. It is important to center and scale your data when
   performing PCA. To do so, use the `scale()` function. Note that scaling and
   centering is done within each column. Given that we want to scale within each
   gene rather than within each sample, you will have to transpose your
   expression matrix prior to scaling then re-transpose it to return to it to
   its original orientation. Since you will have already scaled your data, set
   the variables scale and center in the `prcomp` function equal to
   `FALSE`. Once you have run prcomp, you can view the values for each of the
   principal components by accessing the `$rotation` attribute of your prcomp
   object.
7. Plot PC1 vs PC2 and examine the plot for outliers. You can examine the
   percent variability explained by each principal component by looking at the
   `$importance` attribute.


**Deliverables**:

* comma separated file containing the RMA normalized, ComBat adjusted gene
  expression values
* a histogram of median RLE scores
* a histogram of median NUSE scores
* a plot of PC1 vs PC2 with the percent variability attributed to these
  principal components shown on the x and y axes labels.

In your writeup, provide an interpretation of each of these plots.

## 4. Noise filtering & dimensionality reduction {-}

**Lead Role: Analyst**

To help you get started writing your code, we have provided gene expression
data similar to the form you should obtain in 3.5. You will find a
differential expression result matrix here:

```
/project/bf528/project_1/data/example_intensity_data.csv
```

Microarray analysis is always characterized by the so-called “large *p*, small
*n* ” problem in which the number of features (*p*, e.g. genes) far exceeds the
number of samples (*n*, e.g. patients). Univariate statistical methods, such as
the t-test, are unaffected by the ratio of features to samples. However,
certain multivariate methods, such as clustering, will yield little or no
information when performed in scenarios where $p \gg n$ due to the low
signal-to-noise ratio. There are many basic methods that can be used to remove
this noise. Not all methods are appropriate for every situation, and the choice
of which methods to use and which cutoffs to select must be carefully
considered. Marisa et al. selected genes (probe sets) based on three well
defined metrics. Compute these metrics for your normalized data and use the
cutoffs suggested to arrive at a reduced set of features as described in the
supplemental methods of the paper.

Implement the following filters on the RMA normalized, ComBat adjusted
expression matrix:

1. Expressed in at least 20% of samples (i.e. for each gene, at least 20% of
 the gene-expression values must be > $\log2(15)$).
2. Have a variance significantly different from the median variance of all
   probe sets using a threshold of $p<0.01$ (hint: use a chi-squared test
   as shown in http://www.itl.nist.gov/div898/handbook/eda/section3/eda358.htm.
   You will need to manually compute the test statistic for each gene and
   compare it to the chi-squared distribution with the correct number of
   degrees of freedom using the `qchisq()` function).
3. Have a coefficient of variation > 0.186.
4. Write out a different file containing the gene expression matrix for genes
   passing all three of the filters from 4.1, 4.2, and 4.3.
5. **For groups with Biologist role only:** Write out the expression matrix for
   probesets that pass the expression threshold from 4.2 to a file with
   `write.csv`.

**Deliverables**:

* A comma separated file with the filtered results from all three filters from
  4.4.
* Report the number of genes that pass all of these thresholds.
* **For groups with Biologist role only:** A comma separated file with the
  filtered results from the expression filter from 4.2.

## 5. Hierarchical clustering & subtype discovery {-}

**Lead Role: Analyst**

A powerful analytical tool to leverage with large sample sizes is clustering.
Clustering is an unsupervised method for grouping sets of similar objects based
on some criterion, usually a series of features whose similarity is defined by
some distance function. Cluster analysis is powerful because it doesn’t rely on
a class label (such as disease status), allowing for novel relationships to be
discovered. In this paper, Marisa et al. leveraged a method called Consensus
Clustering to identify the true number of clusters present in their data. This
method is too computationally intensive for the purposes of this course, so we
will use hierarchical clustering instead.

1. Perform hierarchical clustering on your fully filtered data matrix from Part
   4.4. Be sure to check that you are clustering the patients and not the
   genes.
2. Cut the dendrogram such that the samples are divided into 2 clusters. How
   many samples are in each cluster?
3. Create a heatmap of the gene-expression of each gene across all samples
   using the `heatmap()` function. Include a column colorbar by setting the
   `ColSideColors` variable in the heatmap function equal to “red” if the
   sample belongs to the C3 subtype and “blue” otherwise. Subtype annotation
   can be found in the annotation matrix under the title
   `cit-coloncancermolecularsubtype`.
4. Using the expression matrix from Part 4.4 and the cluster memberships from
   Part 5.2, identify genes differentially expressed between the two clusters
   using a Welch t-test (results in a ranked list).  Write out a dataframe
   containing the probeset ID, t-statistic, p-value, and adjusted p-value (i.e.
   FDR, see the `p.adjust` function) to a comma separated file. How many
   genes are differentially expressed at adjusted :math:`p<0.05` between the
   clusters for both lists?
5. Select the most differentially expressed genes that you feel best define the
   clusters and explain your selection.
6. **For groups with Biologist role only**: perform the t-test analysis
   described in 5.4 on the expression matrix from 4.5 as well and provide to
   the Biologist as a csv file.

**Deliverables**:

* report the number of samples in each cluster from Part 5.2
* a heatmap of the genes and samples with a color bar indicating which subtype
  each sample belongs to
* report the number of differentially expressed genes at :math:`p<0.05` between
  the two clusters
* a comma separated file containing the results of the Welch t-test for all
  genes irrespective of significance for each subtype comparison
* report a list of the genes you feel best represent each cluster and explain
  how you came to your conclusion
* **For groups with Biologist role only:** A comma separated file with the
  t-test results computed on the expression matrix from 4.5.


## 6. In-depth Analysis {-}

**Primary role: Biologist - for 4 person groups only**

The authors in Marisa et al sought to understand the biological significance of
the different gene expression profiles for each tumor subtype using gene set
enrichment analysis. Specifically, KEGG, GO, and cancer hallmark genesets were
compared with the top 1000 up- and down-regulated genes of each subtype compared
with all the others using Fisher's Exact test. We will try to reproduce this
analysis using KEGG gene sets and the differential expression results from 5.6.

To help you get started writing your code, we have provided gene expression
statistics similar to the form you should obtain in 5.6. You will find a
differential expression result matrix here:

```
/project/bf528/project_1/data/differential_expression_results.csv
```

1. The differential expression matrix you received has only probeset IDs,
   and not gene symbols. Using the `select()` function of the bioconductor
   package `hgu133plus2.db`, map the probeset IDs to gene symbols by
   specifying the appropriate `key` and `column` arguments. Some probeset
   IDs map to the same gene symbol, so reason about and pick a rationale for
   choosing which probeset ID to use as representative.  Add an additional
   column to the differential expression results that contains one symbol
   for each probeset ID.
2. Locate and download the KEGG, GO, and Hallmark gene sets from [MSigDB](http://software.broadinstitute.org/gsea/msigdb).
   You will need to supply your email address to download the gene sets.
   Download the file with gene symbol identifiers, and not entrez IDs. You
   should have three files with `.gmt` extensions that contain the genesets.
3. Using the differential expression results from your comparisons in 5.6
   that were calculated using the Chi-squared filtered results from 4.5,
   select the top 1000 up- and down-regulated (i.e. positive and negative
   log2 fold change, respectively) genes, irrespective of significance. Create
   a table in your report with the top 10 of these up- and down-regulated genes,
   including t-statistic, p-value, and adjusted p-value. **NB:** You should have
   around ~20k genes listed in the differential expression results from 5.6.
4. We will use the [GSEABase](https://bioconductor.org/packages/release/bioc/html/GSEABase.html)
   bioconductor package to load the genesets we
   downloaded as a GeneSetCollection. Read the package documentation to
   understand how to use the package, and find the function that is used to
   read in GMT formatted files. You may also find our [chapter on gene set
   enrichment analysis](https://bu-bioinfo.github.io/r-for-biological-sciences/biology-bioinformatics.html#gene-set-enrichment-analysis)
   from our course [R for Biological
   Sciences](https://bu-bioinfo.github.io/r-for-biological-sciences/index.html)
   helpful. How many gene sets are there in each of the collections we are
   using?
5. Use the `fisher.test` function to compute hypergeometric statistics and
   p-values comparing overlap for each gene set and each of the top 1000 increased
   and 1000 decreased genes from 6.3. You will have two sets of enrichment statistics,
   one for each direction of effect.

   To implement your test, you might consider writing a function that accepts a set
   of differentially expressed gene symbols and a single gene set to test your code.
   You will need to create a [contingency
   table](https://en.wikipedia.org/wiki/Contingency_table) for each calculation
   to pass as the argument to `fisher.test`, e.g.:

   +-----------------+--------------------------+------------------------------+-------+
   |                 | Differentially Expressed | Not Differentially Expressed | Total |
   +=================+==========================+==============================+=======+
   | In Gene Set     | 26                       | 100                          | 126   |
   +-----------------+--------------------------+------------------------------+-------+
   | Not In Gene Set | 974                      | 18259                        | 19233 |
   +-----------------+--------------------------+------------------------------+-------+
   | Total           | 1000                     | 18359                        | 19359 |
   +-----------------+--------------------------+------------------------------+-------+

   **NB:** These numbers are completely made up. You should use the total
   number of genes in the differential expression results as the number of
   all genes.

   To perform Fisher's Exact test on the above matrix, you could write:

   ```
   > fisher.test(matrix(c(26,974,100,18259),nrow=2))

       Fisher's Exact Test for Count Data

   data:  matrix(c(26, 974, 100, 18259), nrow = 2)
   p-value = 1.119e-09
   alternative hypothesis: true odds ratio is not equal to 1
   95 percent confidence interval:
    3.022027 7.606742
    sample estimates:
    odds ratio
      4.873775
   ```

5. Create a table (dataframe) of statistics for each comparison from 6.2,
   including gene set name, statistic estimate and p-value. Adjust the p-values
   for multiple hypotheses using the Benjamini-Hochberg (FDR) procedure and
   append this adjusted p-value column to the data frame. Finally, sort each
   dataframe by nominal p-value and report the top three results for each in a
   table in your report. Compare the enriched gene sets you found with those
   reported in Figure 2 of the paper.

**Deliverables:**

- a table containing top 10 up- and down-regulated probesets with gene symbol,
  t-statistic, nominal p-value, and adjusted p-value columns
- a description of the gene set databases used, specifying the number of gene
  sets considered in each
- the number of significantly enriched gene sets at adjusted :math:`p<0.05`
- a table containing the top 3 enriched gene sets for each geneset type

## 7. Discuss Your Findings {-}

Discuss your findings with your team members and other teams. Some interesting
questions to consider:

1. If you found a different number of probesets passing the filtering
   thresholds in Part 4, why do you think this is and is this a problem? Does it
   change how you would interpret the results?
2. Which aspects of your analysis, if changed, would have the biggest impact on
   your results?
3. Part 4.2 requires to you to find only genes with variance significantly
   higher than the median variance. Why do you think it is important to consider
   genes with high variance? What information might these gene provide?
4. What biological interpretation may the genes found in Part 4.3 have?

## Assignment Writeup {-}

Refer to the [Project Report Guidelines].
